---
title: "Week 3: Data Visualization"
subtitle: "{ggplot2}"
author: "Eunji Kong <br>(adapted from Dr. Joe Nese's lecture)"
institute: "University of Oregon <br>Fall 2025"
format:
  html:
    toc: true
    toc-depth: 6
    toc-expand: true         
    toc-location: left
    embed-resources: true
    include-in-header: |
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{=html}
<style>
  main.content section.level2 {
    margin-left: 2em;   /* push inward */
    padding-left: 1em;  /* optional spacing */
    border-left: 3px solid #ddd; /* optional vertical line */
  }
  p {
    font-style: italic;
  }
</style>
```


```{r}
#install.packages("tidyverse")
#install.packages("palmerpenguins")
#install.packages("patchwork")
#install.packages("ggridges")
#install.packages("gghighlight")
#install.packages("MetBrewer")
#install.packages("ggthemes")


#library(tidyverse)
#library(palmerpenguins)
#libraryy(patchwork)
#library(ggridges)
#library(gghighlight)
#library(MetBrewer)
#library(ggthemes)
```

# Greetings!

-   Eunji Kong

-   4th year SPED doctoral student

-   Finished EDS specialization

-   EDS project

    -   [Data Viz project](https://eunjikong.shinyapps.io/edld652_shiny/#section-amber-1)
    -   [Capstone](https://eunjikong.shinyapps.io/EDLD640_Capstone_Final/)

# Learning Objectives

-   Understand the basic syntax requirements for {ggplot2}
-   Recognize various options for displaying data
-   Familiarity with various {ggplot2} options/layers
-   **Basically, how to graph and visualize data**

**Lecture/Material Structure**

-   PDF Lecture Notes

    -   Include hyperlinks that take you directly to the relevant topics
        -   Hyperlinks: everything underlined

-   .qmd file (recommended)

    -   Same information as the PDF but allows you to write notes directly in the file

    -   You can also test out code interactively as you follow along

    -   You can then render this document as an HTML file for later review

    -   Visual mode

# `{tidyverse}`

`{tidyverse}` is a a meta-package that loads a set of core packages

```{r message=TRUE}
# If you don't have the package installed
# install.packages("tidyverse")

# load library
library(tidyverse)
```

# `{ggplot2}`

-   gg stands for “grammar of graphics”
-   [**Resources**]{.underline}
    -   ggplot2 book
        -   *email Dr. Nese for digital copy*
    -   Posit [cheat sheet](https://rstudio.github.io/cheatsheets/html/data-visualization.html?_gl=1*1j49xha*_ga*MTM2MTAwMDI2MS4xNjk2OTY0NDAy*_ga_2C0WZ1JHG0*MTY5Njk2NDQwMi4xLjAuMTY5Njk2NDQwMi4wLjAuMA..)
        -   *can be helpful, perhaps more so after a little experience*
    -   [R Graphics Cookbook](http://www.cookbook-r.com/Graphs/)
    -   [R Graph Gallery](https://www.r-graph-gallery.com/index.html)
        -   *past students have really liked this one*

# Components

Every ggplot has **three** components:

1.  **data**
    -   the data used to produce the plot
2.  **aesthetic mappings (aes)**
    -   between variables and visual properties
3.  **layers(s)**
    -   usually through the geom\_\*() function plus various other layers

[**Template**]{.underline}

I use the base R's version of pipe `|>` instead of `%>%` but they are essentially the same thing.

```         
data |> #pipe here
 ggplot(aes(mapping)) + #plus here
 geom_function() +
 additional layers
```

Above code is the same as the bottom code.

```         
ggplot(data, aes(mapping)) +
geom_function() +
additional layers
```

## data

```{r}

# install.packages("palmerpenguins")

library(palmerpenguins)
```

```{r}

head(penguins)

# str(penguins)

# glimpse(penguins)

# colnames(penguins)

# View(penguins)
```

------------------------------------------------------------------------

## ggplot(aes(mapping))

-   aesthetic mappings describe how **variables** in the data are **mapped to visual properties**

-   Some visual properties include:

    -   **x**

    -   **y**

    -   color [*(will come back to it)*]{style="color:darkgreen;"}

    -   fill [*(will come back to it)*]{style="color:darkgreen;"}

    -   alpha [*(will come back to it)*]{style="color:darkgreen;"}

    -   others (linetype, shape, linewidth, size, group)

```{r}
penguins |> 
  ggplot(aes(x = bill_length_mm, y = body_mass_g))
```

**QUESTION: What do you see? Why is there nothing plotted?**

```         
ANSWER: 
```

------------------------------------------------------------------------

## Layers

### geom_function()

Use a geom_function() to represent **data points**

+---------------------------+---------------------+---------------------------+---------------------------------------+
|                           | **Only 1 Variable** | **Continuous Variable 2** | **Discrete Variable 2**               |
+---------------------------+---------------------+---------------------------+---------------------------------------+
| **Continuous Variable 1** | geom_histogram      | geom_point                | geom_density_ridges (from {ggridges}) |
|                           |                     |                           |                                       |
|                           | geom_density        | geom_smooth               | geom_boxplot                          |
|                           |                     |                           |                                       |
|                           |                     | geom_line                 | geom_violin                           |
|                           |                     |                           |                                       |
|                           |                     |                           | geom_col                              |
+---------------------------+---------------------+---------------------------+---------------------------------------+
| **Discrete Variable 1**   | geom_bar            | x                         | geom_count                            |
+---------------------------+---------------------+---------------------------+---------------------------------------+

**Other**

Heatmap: geom_tile

------------------------------------------------------------------------

#### geom_histogram() {#geom_histogram}

**General research question**: How does the values of my continuous variable vary across its range?

**Our data specific question**: How is the distribution of penguin bill lengths (mm) in this sample? Any outliers? Unimodal?

```{r}

penguins |> 
  ggplot(aes(x = bill_length_mm)) + # Remember to use + instead of |> or %>%
  geom_histogram()

```

##### Color vs Fill

```{r}
penguins |> 
  ggplot(aes(x = bill_length_mm)) +
  geom_histogram(color = "blue")

```

```{r}
penguins |> 
  ggplot(aes(x = bill_length_mm)) +
  geom_histogram(color = "blue", 
                 fill = "green")
```

Color = outline

Fill = area

##### Transparency

```{r transparency}
penguins |> 
  ggplot(aes(x = bill_length_mm)) +
  geom_histogram(color = "blue", 
                 fill = "green",
                 alpha = 0.2)

```

Color, fill & alpha in this example area all [**fixed**]{.underline} settings (i.e., applies to all data points).

##### More aes mapping

```{r}
penguins |> 
  ggplot(aes(x = bill_length_mm)) +
  geom_histogram(aes(fill = sex), # note that fill is inside  aes()
                 alpha = 0.7)

```

Fill here is a [**conditional**]{.underline} mapping, meaning that the fill color is different based on the variable (in this case the sex of the birds).

##### Fixed vs Conditional

```         
penguins |> 
ggplot(aes(x = bill_length_mm)) +
geom_histogram(fill = "green")
```

In the above example where fill is not within aes(), fill is a [**fixed**]{.underline} setting. Also notice that color is in quotes.

```         
penguins |> 
ggplot(aes(x = bill_length_mm)) +
geom_histogram(aes(fill = sex))
```

In the above example, aes() is used to access variables and make changes according to a specific variable. Here, fill is a [**conditional**]{.underline} on the variable, sex. Also notice that variables are not in quotes.

```{r}
a <- penguins |> 
ggplot(aes(x = bill_length_mm)) +
geom_histogram(fill = "green")

b <- penguins |> 
ggplot(aes(x = bill_length_mm)) +
geom_histogram(aes(fill = sex))

#install.packages("patchwork")
library(patchwork)
a+b


```

**Be mindful of aes()**

```         
penguins |>
ggplot(aes(x = bill_length_mm))+
geom_histogram(fill = “green”)
```

```         
penguins |>
ggplot(aes(x = bill_length_mm))+
geom_histogram(aes(fill = “green”))
```

**Question**: What is wrong with the bottom code? How do you think the plot will look like?

```         
Answer: 
```

```{r}
penguins |> 
ggplot(aes(x = bill_length_mm)) +
geom_histogram(aes(fill = "green"))
```

#### geom_density()

**General research question:** How does the probability density of my continuous variable vary across its range?

Think of it as a smoothed histogram

-   **Difference**: not use bins; not use count but use relative frequency per unit of x

```{r}

penguins |>
  ggplot(aes(x = bill_length_mm)) + 
  geom_density()

```

##### More aes mapping

```{r}

penguins |>
  ggplot(aes(x = bill_length_mm)) +
  geom_density(aes(fill = sex))

```

Add transparency for clarity

```{r}
penguins |>
  ggplot(aes(x = bill_length_mm)) +
  geom_density(aes(fill = sex), alpha = 0.5)
```

##### Histogram vs Density

```{r}
a <- penguins |>
  ggplot(aes(x = bill_length_mm)) +
  geom_histogram(aes(fill = sex), alpha = 0.5)

b <- penguins |>
  ggplot(aes(x = bill_length_mm)) +
  geom_density(aes(fill = sex), alpha = 0.5)

a + b
```

**Question**: What is the difference that you see? When would you use one vs another?

```         
Answer: 
```

##### facet_wrap

wrap by 1 variable

```{r}
penguins |>
  ggplot(aes(x = bill_length_mm)) +
  geom_density(aes(fill = sex), alpha = 0.5) +
  facet_wrap(~sex) # remember to use ~
```

wrap by 2 variables

```{r}
penguins |>
  ggplot(aes(x = bill_length_mm)) +
  geom_density(aes(fill = sex), alpha = 0.5) +
  facet_wrap(year~sex)
```

wrap using vars()

```{r}
penguins |>
  ggplot(aes(x = bill_length_mm)) +
  geom_density(aes(fill = sex), alpha = 0.5) +
  facet_wrap(vars(year,sex))
```

#### geom_density_ridges()

geom_density_ridges: two variables

```{r}
# install.packages("ggridges")

library(ggridges)

penguins |>
  ggplot(aes(bill_length_mm, sex)) +
  geom_density_ridges()
```

#### geom_point()

**General research question**: How are two numeric variables related? (raw observations)

**Our data specific question**: What is the relationship between penguin’s bill length and body mass?

```{r}
penguins |>
  ggplot(aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point()

```

**Add color**

```{r}
penguins |>
  ggplot(aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point(color = "magenta")

```

**Emphasize specific data points (island = Torgersen)**

```{r}
penguins |>
  ggplot(aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point(color = "magenta") +
  geom_point(data = filter(penguins, island == "Torgersen"), color = "blue")
```

```{r}
penguins |>
  ggplot(aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point(data = filter(penguins, island == "Torgersen"), color = "blue") +
  geom_point(color = "magenta")
  
```

**Question**: What happened when we switched the order of the geom_points?

```         
Answer: 
```

**Emphasize another way**

```{r}
# install.packages("gghighlight")



penguins |>
  ggplot(aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point(color = "magenta") +
  gghighlight::gghighlight(island == "Torgersen")
```

#### geom_smooth()

**General research question**: What is the pattern of relationship of two continuous variables? (trend)

**Our data specific question**: What is the trend or pattern of relationship between penguin’s bill length and body mass?

```{r message=TRUE}
penguins |>
  ggplot(aes(bill_length_mm, body_mass_g)) + 
  geom_smooth()
```

##### Method

```{r}
penguins |>
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_smooth(method = "lm")
```

No need to include “x =” or “y =” because ggplot assumes the first argument will be x and then y.

```{r}
penguins |>
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_smooth(method = "lm", level = .65)
```

```{r}
penguins |>
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_smooth(method = "lm", se=FALSE)
```

**Note**: This is not the same as geom_line(). We are fitting a line of best fit with geom_smooth()

**Adding Layers**

```{r}
penguins |>
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm")
```

##### Global

If we use something like color = “x” in the first aesthetic, it will carry on through all additional layers.

```{r}
penguins |> 
  ggplot(aes(bill_length_mm, body_mass_g, color = species)) + #color = spieces
  geom_point() +
  geom_smooth(method = "lm")
```

```{r}
penguins |> 
  ggplot(aes(bill_length_mm, body_mass_g, color = species)) +
  geom_point(aes(color = species)) +
  geom_smooth(method = "lm", aes(color = species))
```

##### Local

```{r}
penguins |> 
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_point(aes(color = species)) + #color = spieces
  geom_smooth(method = "lm")
```

#### geom_line()

-   geom_point: raw observations, not linked

-   geom_smooth: trend/pattern

-   geom_line: raw data linked

```{r}
penguins |>
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm")
```

```{r}
penguins |>
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  geom_line()
```

When should you use line plots?

-   Usually when time is involved

-   One time point per line or per group

-   Shows linkage

```{r}
# Original data
head(penguins)

```

```{r}
# Create new data set so that there is only one data point per year
penguins_year <- penguins |>
  group_by(year) |> 
  summarize(avg_bill = mean(bill_length_mm, na.rm=TRUE))
head(penguins_year)
```

```{r}
penguins_year |>
  ggplot(aes(year, avg_bill)) + 
  geom_line()
```

```{r}
# Create new data set so that there is one data point for each year for each species
penguins_year_species <- penguins |> 
  group_by(year, island) |> 
  summarize(avg_bill = mean(bill_length_mm, na.rm=TRUE))

head(penguins_year_species)
```

```{r}
penguins_year_species |>
  ggplot(aes(year, avg_bill, group = island, color = island)) + 
  geom_line()

```

#### geom_boxplot()

**General research question**: How is a continuous variable distributed across groups, and how do the medians, quartiles, and potential outliers compare?

```{r}
penguins |>
  ggplot(aes(species, body_mass_g)) +
  geom_boxplot()
```

#### geom_violin()

**General research question**: How is the full distribution of a continuous variable shaped across groups?

```{r}
penguins |>
  ggplot(aes(species, body_mass_g)) +
  geom_violin()
```

#### geom_bar()

##### geom_bar() vs geom_col()

+----------------------------------------------------------------------------+------------------------------------------------------------------------------------+
| geom_bar()                                                                 | geom_col()                                                                         |
+============================================================================+====================================================================================+
| -   1 discrete variable                                                    | -   2 variables                                                                    |
|                                                                            |                                                                                    |
|                                                                            |     -   1 continuous                                                               |
|                                                                            |                                                                                    |
|                                                                            |     -   1 discrete (unique)                                                        |
+----------------------------------------------------------------------------+------------------------------------------------------------------------------------+
| -   counts rows                                                            | -   need to have a variable with numbers in your data (average, proportion, count) |
|                                                                            |                                                                                    |
| -   height of the bar is proportional to the number of cases in each group |                                                                                    |
+----------------------------------------------------------------------------+------------------------------------------------------------------------------------+

```{r}
penguins |> 
  ggplot(aes(species)) + # one variable in the `aes()`
  geom_bar()
```

#### geom_col()

```{r}
summarized_penguins <- penguins |> 
  group_by(species) |> 
  summarize(N = n())

head(summarized_penguins)
```

```{r}
summarized_penguins |>
  ggplot(aes(species, N)) +
  geom_col()
```

##### More aes mapping

```{r}
summarized_penguins2 <- penguins |>
  group_by(species, sex) |>
  na.omit() |> 
  summarize(bill_length_avg = mean(bill_length_mm))

summarized_penguins2

```

```{r}
ggplot(summarized_penguins2, aes(species, bill_length_avg)) +
  geom_col(aes(fill = sex))

```

##### Position

```{r}
ggplot(summarized_penguins2, aes(species, bill_length_avg)) +
  geom_col(aes(fill = sex), position = "dodge")

```

##### coord_flip

```{r}
ggplot(summarized_penguins2, aes(species, bill_length_avg)) +
  geom_col(aes(fill = sex), position = "dodge") +
  coord_flip()
```

#### geom_count()

**General research question**: How many observations fall in each category pair?

**Our data specific question**: How many of each species live in each island?

```{r}
penguins |>
  ggplot(aes(species, island)) +
  geom_count()

```

##### Scales

```{r}
penguins |>
  ggplot(aes(species, island)) +
  geom_count(aes(color = after_stat(n)))+
  scale_color_gradient(low = "lightblue", high = "brown")
```

**What do scales do?**

Scales control how the mappings you added to aes are displayed (e.g., color range, size range, breaks and labels, range or limits)

Template: scale\_\*

Most aes mappings: x, y, size, color, fill, line, alpha, etc

**Colorblind friendly**

```{r}
penguins |>
  ggplot(aes(species, island)) +
  geom_count(aes(color = after_stat(n)))+
  scale_color_viridis_c()
```

```{r}
penguins |>
  ggplot(aes(species, island)) +
  geom_count(aes(color = after_stat(n)))+
  scale_color_viridis_c(option = "turbo") #magma, interno, plasma, viridis, cividis, rocket, mako, turbo or A-H
```

-   [{MetBrewer}](https://github.com/BlakeRMills/MetBrewer) - inspired by art in the Met

```{r MetBrewer}
#install.packages("MetBrewer")

penguins |>
  ggplot(aes(species, island)) +
  geom_count(aes(color = after_stat(n)))+
  scale_color_gradientn(colors=MetBrewer::met.brewer("Isfahan1"))

```

#### geom_tile()

**General research question**: What’s the value of a numerical measure (Z) for each (X, Y) pair? In other words, what is the correlation (Z) for each X,Y pair?

```{r}
corr <- penguins |>
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) |>
  drop_na() |>                
  cor()

pc <- corr |> 
  as.data.frame() |> 
  rownames_to_column(var = "row") |> 
  pivot_longer(
    cols = -row,
    names_to = "col",
    values_to = "cor")

head(pc)

```

```{r}
ggplot(pc, aes(row, col, fill = cor)) +
  geom_tile()
```

```{r}
ggplot(pc, aes(row, col, fill = cor)) +
  geom_tile() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
ggplot(pc, aes(row, col, fill = cor)) +
  geom_tile() +
  scale_fill_viridis_c()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

------------------------------------------------------------------------

### Other Layers

#### labels

##### axis labels

```{r}
penguins |> 
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_point(aes(color = species)) 
```

```{r}
penguins |> 
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_point(aes(color = species)) + 
  labs(x="Bill length (mm)",
       y="Body mass (g)")

```

##### title

```{r}
penguins |> 
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_point(aes(color = species)) + 
  labs(x="Bill length (mm)",
       y="Body mass (g)",
       title = "Relationship between bill length and body mass")

```

##### subtitle

```{r}
penguins |> 
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_point(aes(color = species)) + 
  labs(x="Bill length (mm)",
       y="Body mass (g)",
       title = "Relationship between bill length and body mass",
       subtitle="Grouped by species")

```

##### caption

```{r}
penguins |> 
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_point(aes(color = species)) + 
  labs(x="Bill length (mm)",
       y="Body mass (g)",
       title = "Relationship between bill length and body mass",
       subtitle="Grouped by species",
       caption = "palmerpenguins")

```

##### tag

```{r}
penguins |> 
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_point(aes(color = species)) + 
  labs(x="Bill length (mm)",
       y="Body mass (g)",
       title = "Relationship between bill length and body mass",
       subtitle="Grouped by species",
       caption = "palmerpenguins",
       tag = "(A)")

```

##### legend (one way)

```{r}
penguins |> 
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_point(aes(color = species)) + 
  labs(x="Bill length (mm)",
       y="Body mass (g)",
       title = "Relationship between bill length and body mass",
       subtitle="Grouped by species",
       caption = "palmerpenguins",
       tag = "(A)",
       color="SPECIES!")

```

------------------------------------------------------------------------

#### theme

The default is theme_gray(). There are a lot of built-in alternative in {ggplot2}. My go-to is theme_minimal() because it is clean without a lot of unnecessary visuals.

If you want to set theme globally (meaning to all your graphs in your document), add theme_set(theme_minimal()) to the first line after you load your libraries.

```{r}
penguins |> 
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_point(aes(color = species)) + 
  labs(x="Bill length (mm)",
       y="Body mass (g)",
       title = "Relationship between bill length and body mass",
       subtitle="Grouped by species",
       caption = "palmerpenguins",
       tag = "(A)",
       color="SPECIES!") +
  theme_minimal()
```

Other packages:

-   [{ggthemes}](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/)

-   [{ggthemeassist}](https://github.com/calligross/ggthemeassist)

-   [{hrbrthemes}](https://github.com/hrbrmstr/hrbrthemes)

-   [Build your own theme](https://bookdown.org/rdpeng/RProgDA/building-a-new-theme.html)

```{r }
#install.packages("ggthemes")

penguins |> 
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_point(aes(color=species)) + 
	ggthemes::theme_economist()+
  ggthemes::scale_color_economist()

```

```{r build-your-own-theme}
penguins |> 
  ggplot(aes(bill_length_mm, body_mass_g)) +
  geom_point(aes(color = species)) + 
  labs(x        = "Bill length (mm)",
       y        = "Body mass (g)",
       title    = "Relationship between bill length and body mass",
       subtitle = "Grouped by species",
       caption  = "palmerpenguins",
       tag      = "(A)",
       color    = "SPECIES!") +
  theme(plot.title        = element_text(size=13, face="bold", hjust =0.5), 
        axis.title        = element_text(size=11, family="Georgia"), 
        axis.text.x       = element_text(size=10, angle = 45, hjust=1),
        panel.background  = element_rect(fill = "grey95"),
        plot.background   = element_rect(fill = "white"),
        panel.grid.major  = element_line(color = "black"),
        panel.grid.minor  = element_blank(),
        legend.position   = "top",
        legend.title      = element_text(face="bold"),
        legend.background = element_rect(fill = "transparent")) 

```

------------------------------------------------------------------------

# Practice together

## 1

Get to know the data - str(mpg) or head(mpg)

```{r}

```

## 2

What is the overall distribution of city fuel efficiency (mpg) across car models?

```{r}

```

## 3

How does the distribution vary by drivetrain type (e.g., front-, rear-, 4-wheel drive)?

```{r}

```

## 4

What is the relationship between city and highway mpg?

```{r}

```

## 5

Can we focus on/emphasize Audi’s relationship?

```{r}

```

## 6

Can we have larger points for clarity?

```{r}

```

## 7

How are the city/hwy mpg relationships different by car class?

```{r}

```

## 8

Too much clutter. Can we just see trends?

```{r}

```

## 9

Still too much clutter. Better way to clearly see each trends?

```{r}

```

## 10

Can we make it colorblind friendly?

```{r}

```

## 11

Can we clarify axis and legend labels?

```{r}

```

## 12

Can we polish the appearance with a theme?

```{r}

```

